JD notes on effects

### 2022-06-14 manuscript/abstract.tex
Adopting th notion of predictors as opposed to model variables

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Describing the curves: uncertainty estimation and bias correction for predictor effects in simple and generalized linear (mixed) models} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Steve Cygu\textsuperscript{1},
Benjamin M. Bolker\textsuperscript{1,2},
Jonathan Dushoff\textsuperscript{1,2}
\\
\bigskip
\textbf{1} School of Computational Science and Engineering, McMaster University, Hamilton, Ontario, Canada
\\
\textbf{2} Department of Biology, McMaster University, Hamilton, Ontario, Canada
\\
\bigskip

@@

In applications that use generalized linear (mixed) models, outcome predictions are often of interest. For models that involve complex multiplicative interactions, additional non-focal predictors or nonlinear link functions, the estimated coefficients are not readily interpretable. A general way to summarize these kind of models is through predictions, that depend not only on choice of representative values of focal predictor but are also sensitive to which values of non-focal predictors are chosen. The most common approach is generating predictions at a reference point, usually the mean, of non-focal predictor, i.e., mean-based (centered) which could be considered as the effect of an average case in the population. In the presence of additional non-focal predictors, nonlinear link functions, random effect terms, etc., mean-based approach generate predictions that are biased and may not be consistent with the observed quantities. An alternative is the whole-sample-based approach which estimates the average effect in the population. Moreover, anchored effects provide an alternative and a more clear way to describe uncertainty associated with the focal predictor of interest. In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that whole-sample-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction. We also present an alternative way, anchored confidence intervals, to describe uncertainties associated with these predictions.

\section*{Definitions}

\begin{itemize}
\item \textbf{Input variables}: Scientific variables underlying an inference or exploration. The focal predictor we use for effects or predictions is an input variable.
\item \textbf{Model variables}: Variables that represent columns in the model matrix. Each input variable will correspond to one or more model variables. In particular, variables with more than two categories, or variables modeled with a spline or polynomial response, will correspond to more than one model variables.
\item\textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables). The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable.
\item \textbf{Reference point}: Value or values chosen for \emph{non-focal predictors}, when estimating effects. Typically the center point, but can instead be a population of quantiles or observations. 
\item \textbf{Anchor}: The value chosen for the \emph{focal predictor} when estimating effect confidence intervals (anchor choice does not affect the estimate). Typically chosen as the center point of the model variables corresponding to the focal input variable.
\end{itemize}

>>>>

% Title must be 250 characters or less.
\begin{flushleft}
{\Large
\textbf\newline{Describing the curves: uncertainty estimation and bias correction for predictions in simple and generalized linear (mixed) models} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
}
\newline
% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
\\
Steve Cygu\textsuperscript{1},
Benjamin M. Bolker\textsuperscript{1,2},
Jonathan Dushoff\textsuperscript{1,2}
\\
\bigskip
\textbf{1} School of Computational Science and Engineering, McMaster University, Hamilton, Ontario, Canada
\\
\textbf{2} Department of Biology, McMaster University, Hamilton, Ontario, Canada
\\
\bigskip

@@

In applications that use generalized linear (mixed) models, outcome predictions are often of interest. For models that involve complex multiplicative interactions, additional non-focal input variables or nonlinear link functions, the estimated coefficients are not readily interpretable. A general way to summarize these kind of models is through predictions, that depend not only on choice of representative values of focal input variables but are also sensitive to which values of non-focal input variables are chosen. The most common approach is generating predictions at a reference point, usually the mean, of non-focal predictor, i.e., mean-based (centered) which could be considered as the effect of an average case in the population. In the presence of additional non-focal input variables, nonlinear link functions, random effect terms, etc., mean-based approach generate predictions that are biased and may not be consistent with the observed quantities. An alternative is the whole-sample-based approach which estimates the average effect in the population. Moreover, anchored effects provide an alternative and a more clear way to describe uncertainty associated with the focal input variable. In addition to theoretical and methodical comparison, using simulation, we illustrate the two approaches and show that they can produce substantially different results and that whole-sample-based approach can not only produce estimates consistent with the observed values, but also appropriate for bias correction. We also present an alternative way, anchored confidence intervals, to describe uncertainties associated with these predictions.

### 2022-06-14 manuscript/body.tex
Adopting th notion of predictors as opposed to model variables

\jd{How did all the figures wind up in back?}

@@

In applications that use regression models, outcome predictions are often of interest. Predictions provide a great way to summarize what the regression model is telling us and very useful for interpreting and visualizing model estimates. For example, in logistic regression models, the coefficient estimates, when reported as log odds, are not easy to interpret and are less informative. Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models interactions or other nonlinear transformations, the magnitude of the effect of change in the outcome depends on the values of the predictor of interest and other predictors. Hence, the conclusions one can make about the estimated effects greatly depends on how well we choose the values of the other predictors, i.e., \emph{non-focal} predictors.

@@

\section{Quantities of interest}

\jd{Is this borrowed from me? I'm not really following the structure of the MS right now. Maybe I'll be clearer in the morning.}

Several quantities of interest may be derived from regression models. The first one is the coefficient estimates. Others are \emph{input and model variables}, \emph{model matrix}, \emph{model center}, \emph{reference point} and \emph{anchor}.

\subsection{Input variables}

These are scientific variables underlying an inference or exploration. The focal and non-focal predictors we use for effects or predictions are an input variables.

\subsection{Model matrix}

Model matrix refers to the design matrix whose rows include all combination of variables appearing in the interaction terms, along with the typical values of the focal and non-focal predictors.

\subsection{Model variables}

These are variables that represent columns in the model matrix. Each input variable may correspond to one or more model variables. In particular, variables with more than two categories, or variables modeled with a spline or polynomial response, will correspond to more than one model variables.

\subsection{Model center}

A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables). We also refer to this as mean-based approach. The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in Fig~\ref{fig:justify_plots}. In simple linear models (without) interaction, the input variable can be used as a proxy.

\subsection{Reference point}

Corresponds to the value or values chosen for non-focal predictors, when estimating the predictions. Typically the center point (mean-based), but can instead be a whole-sample of quantiles or observations. Appropriately choosing the reference point is very crucial in generating ``correct'' predictions especially when dealing with non-linear link functions.

\subsection{Anchor}

The value chosen for the focal predictor when estimating effect confidence intervals (anchor choice does not affect the main prediction estimates). Typically chosen as the center point of the model variables corresponding to the focal input variable but other sensible values can be chosen too.

If we are trying to focus on effects, the anchor from which we calculate CIs becomes important. This is not true for the more conventional prediction plots: we can choose any anchor, e.g., zero-anchored, but model center seems like a natural, stable choice, i.e., center-anchored. In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the minimum and center (mean) of the focal predictor.

>>>>

@@

In applications that use regression models, outcome predictions are often of interest. Predictions provide a great way to summarize what the regression model is telling us and very useful for interpreting and visualizing model estimates. For example, in logistic regression models, the coefficient estimates, when reported as log odds, are not easy to interpret and are less informative. Since logistic models are nonlinear (due to the nonlinear link function), in multivariate models interactions or other nonlinear transformations, the magnitude of the effect of change in the outcome depends on the values of the input variable of interest and the other remaining input variables. Hence, the conclusions one can make about the estimated predictions greatly depends on how well we choose the values of the other input variables, i.e., \emph{non-focal} input variables.

@@

\section{Definitions}

In order to discuss more general statistical background and mathematical formulation, we need to understand and formally define a number of terms, some of which have been introduced in the previous section.

\begin{itemize}
\item \textbf{Predictors}: Refer to the X-variables which go into the regression model. For example, consider the model the includes the interaction in Equation~\ref{eq:simple_inter_higher}. This regression model has $5$ predictors -- $x_1, x_2, x_3, x_2x_3$ and the intercept term.
\item \textbf{Focal and non-focal predictors}: Focal predictor refers to the predictor we are interested in generating its effects on the predictions. For multivariate models, all other remaining predictors are non-focal predictors. For example, consider model described by Equation~\ref{eq:simple_inter_higher_no_interaction}. Suppose we are interested in the effect of $x_1$ on the predicted $y$, then $x_1$ is the focal predictor while $x_2$ and $x_3$ are non-focal predictors.
\item \textbf{Input variables}: Refers to the observed (or scientific) variables underlying an inference or exploration, and forms the X-variables. For example, the regression model described by Equation~\ref{eq:simple_inter_higher} has only $4$ input variables -- $x_1, x_2, x_3$.
\item \textbf{Model matrix}: Refers to the design matrix whose rows include all combination of variables appearing in the interaction terms. Consider an example for three hypothetical households -- the first household head is a Christian with an income of $\$ 50$, the second household head is Muslim with an income of $\$ 100$ while the third household head is a Jew with an income of $\$ 77$. Suppose we want to model the household size (\code{hhsize}) as function of these household characteristics, i.e., $$\mathrm{hhsize} = \beta_0 + \beta_1\times\mathrm{income} + \beta_{2[r]}\times\mathrm{religion} + \mathrm{error}.$$ The model matrix corresponding to this model is given by
$$\begin{bmatrix}{}
 Intercept & income & religionJew & religionMuslim \\
 1 & 50 & 0 & 0 \\
  1 & 100 & 0 & 1 \\
  1 & 77 & 1 & 0 \\
\end{bmatrix}.$$ The first column represents the constant term in our model, $\beta_0$. For continuous variables, the representation in the model matrix is the same as the correspond input variables (for example second column representing income). For categorical variables, however, by default, the model matrix  creates additional dummy variables using the reference cell parameterization. This means that, if an input variable has $L$ factor levels, then there will be $L-1$ dummy columns representing all but the first level created in the model matrix. In our example, the column \code{religion} is missing and instead we have \code{religionJew} and \code{religionMuslim}; the missing category \code{religionChristian} is treated as the reference category.
\item \textbf{Model variables:} It is easy to interchangeably refer to predictors as input variables, and vice versa. To avoid this issue, we adopt the term model variables which are variables that represent columns in the model matrix. Each input variable may correspond to one or more model variables. In particular, variables with more than two categories (for instance religion in our previous example), or variables modeled with a spline or polynomial term, will correspond to more than one model variables.
\item \textbf{Model center:} A point corresponding to a column-wise mean of the model matrix (the mean of one or more model variables). Also referred to as center point. We refer to the prediction approach which uses model center as mean-based. The center point for a set of model variables corresponding to an input variable may not represent a possible value of the input variable -- the case in Fig~\ref{fig:justify_plots}. In simple linear models (without) interaction, the mean of the input variable can be used as a proxy for center point.
\item \textbf{Reference point:} Corresponds to the value or values chosen for non-focal predictors, when estimating the predictions. Typically the center point (mean-based), but can instead be a whole-sample of quantiles or observations. Appropriately choosing the reference point is very crucial in generating correct predictions especially when dealing with non-linear link functions.
\item \textbf{Anchor:} The value chosen for the focal predictor when estimating effect confidence intervals (anchor choice does not affect the main prediction estimates). Typically chosen as the center point of the model variables corresponding to the focal input variable but other sensible values can be chosen too. If we are trying to focus on effects, the anchor from which we calculate CIs becomes important. This is not true for the more conventional prediction plots: we can choose any anchor, e.g., zero-anchored, but model center seems like a natural, stable choice, i.e., center-anchored. In Fig~\ref{fig:justify_anchors}, we compare effects plots anchored at the minimum and center (mean) of the focal predictor.
\end{itemize}
